---
title: "Dyslexia Identification: Texas Legislative Trends in Prevalence Rate of Students by School District Locale"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc_float: true
    code_folding: show
    self_contained: false
---

This webpage contains outputs and R codes used for the two-level multilevel modeling in Authors (2024). Data analysis scripts have been posted through an online data repository, accessible at <https://osf.io/gqpfk/?view_only=32be323dc9a6471c9bd721415c24debc>.


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  message = FALSE, 
  warning = FALSE)
```

# Data Analysis {.tabset}

## Preprocess 

#### Load libraries.

```{r}
suppressPackageStartupMessages({
    library(tidyverse)
    library(dplyr)
    library(fastDummies)
    library(nlme)
    library(lmeInfo)
    library(clubSandwich)
    library(plotly)
    library(ggplot2)
})

load("output.RData")
```

#### Preprocess data.

```{r, eval = FALSE}
data = read.csv("data.csv") 

data$NCES_district_locale <- gsub("-.*", "", data$NCES_district_locale)

data <- dummy_cols(data, select_columns = "NCES_district_locale", remove_selected_columns = FALSE)
colnames(data) <-  gsub("^NCES_district_locale_", "", colnames(data))

data <- data %>% 
    mutate(prevalence_rate = round((dyslexia_n / total_n) * 100, 2)) %>%
    mutate(year_2017 = year - 2017) %>%
    mutate(slope_2017 = year_2017) %>%
    mutate(year_2018 = year - 2018) %>%
    mutate(knot_2018 = as.integer(I(year >= 2018 ))) %>%
    mutate(slope_change_2018 = year_2018 * knot_2018) %>%
    mutate(year_2019 = year - 2019) %>%
    mutate(knot_2019 = as.integer(I(year >= 2019 ))) %>%
    mutate(slope_change_2019 = year_2019 * knot_2019) %>%
    mutate(year_2020 = year - 2020) %>%
    mutate(knot_2020 = as.integer(I(year >= 2020 ))) %>%
    mutate(slope_change_2020 = year_2020 * knot_2020) %>%
    mutate(year_2021 = year - 2021) %>%
    mutate(knot_2021 = as.integer(I(year >= 2021 ))) %>%
    mutate(slope_change_2021 = year_2021 * knot_2021) %>%
    mutate(year_2022 = year - 2022) %>%
    mutate(knot_2022 = as.integer(I(year >= 2022 ))) %>%
    mutate(slope_change_2022 = year_2022 * knot_2022) 
```

## Visualize 

```{r, echo = FALSE, eval = FALSE}
geom_line_plot <- data %>%
  group_by(NCES_district_locale, school_year) %>%
  summarize(rate = mean(prevalence_rate)) %>%
  ungroup() %>%
  ggplot(aes(x = school_year, y = rate, group = NCES_district_locale, color = NCES_district_locale)) +
  geom_line() +
  geom_line(data = data %>%
              group_by(school_year) %>%
              summarize(rate = mean(prevalence_rate)),
            aes(x = school_year, y = rate, group = 1), 
            color = "black", size = 1.5, linetype = "solid") +
  labs(title = "",
       x = "School Year",
       y = "Prevalence Rate (%)") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 11),
    legend.position ="top",
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 11),
    legend.title.align = 0.5,
    axis.line = element_line(),          
    strip.text.x = element_text(color = "#3B3B3B", size = 11),
    axis.line.x = element_line(color = "#3B3B3B"),
    axis.text.x = element_text(size = 11, color = "#3B3B3B", angle = 45, hjust = 1),
    axis.ticks = element_line(color = "#3B3B3B"),
    axis.ticks.length = unit(0.1, "cm"),  
    axis.text.y = element_text(size = 11, color = "#3B3B3B"),
    axis.title = element_text(face = "bold", size = 11, color = "#3B3B3B")
  ) + 
  labs(color = "District Locale") 
```


```{r, echo = FALSE}
geom_line_plot %>% ggplotly()
```

```{r, echo = FALSE, eval = FALSE}
geom_col_plot <- data %>%
  group_by(NCES_district_locale, school_year) %>%
  summarize(rate = mean(prevalence_rate)) %>%
  ggplot(aes(x = school_year, y = rate, group = NCES_district_locale, color = NCES_district_locale)) +
  geom_col(aes(fill = NCES_district_locale)) + 
  labs(title = "",
       x = "School Year",
       y = "Prevalence Rate (%)") +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 11),
    legend.position ="top",
    legend.text = element_text(size = 11),
    legend.title = element_text(size = 11),
    legend.title.align = 0.5,
    axis.line = element_line(),          
    strip.text.x = element_text(color = "#3B3B3B", size = 11),
    axis.line.x = element_line(color = "#3B3B3B"),
    axis.text.x = element_text(size = 11, color = "#3B3B3B", angle = 45, hjust = 1),
    axis.ticks = element_line(color = "#3B3B3B"),
    axis.ticks.length = unit(0.1, "cm"),  
    axis.text.y = element_text(size = 11, color = "#3B3B3B"),
    axis.title = element_text(face = "bold", size = 11, color = "#3B3B3B")
  ) +
  labs(fill = "District Locale", color = "District Locale")
```


```{r, echo = FALSE}
geom_col_plot %>% ggplotly()
```

## Null Model

#### Check residuals for normality

```{r, eval=FALSE}
Model.0 <- lme(
    prevalence_rate ~ 1, 
    data = data,
    method = "REML",
    random = ~ 1 | district,
    control = list(
        maxIter = 100, 
        msMaxIter = 100, 
        tolerance = 1e-3, 
        opt = "optim",
        optimMethod = "BFGS"
    )
)
```

#### Cluster-robust variance estimation

```{r}
options(width = 100)
Model.0.vcov <- vcovCR(Model.0, type = "CR2") # Calculate variance-covariance matrix
Model.0.crve <-
  coef_test(Model.0, vcov = Model.0.vcov , test = "Satterthwaite") # Test regression coefficients 
Model.0.crve
```

#### Variance components

```{r}
VarCorr(Model.0, which = "var-cov")
```

#### Standard errors of variances

```{r}
sqrt(diag(varcomp_vcov(Model.0)))
```
####  Intraclass correlation coefficient (ICC)

```{r}
total_var <- as.numeric(VarCorr(Model.0)[[2]]) +
  as.numeric(VarCorr(Model.0)[[1]]) 

ICC.L2 = as.numeric(VarCorr(Model.0)[[1]]) / total_var
ICC.L2
```

## Model 1

#### Legislative trends in the prevalence rate of students identified as having characteristics of dyslexia in Texas school districts.

```{r, eval = FALSE}
Model.1 <- lme(
  prevalence_rate ~ slope_2017 + slope_change_2018 + slope_change_2019 + 
    slope_change_2020 + slope_change_2021 + slope_change_2022, 
  data = data,
  method = "REML",
  random = ~ 1 | district,
  correlation = corAR1(value = 0.2, form = ~ year | district),
  control = list(
    maxIter = 100, 
    msMaxIter = 100, 
    tolerance = 1e-3, 
    opt = "optim",
    optimMethod = "BFGS"
  )
)
```

#### Cluster-robust variance estimation

```{r}
options(width = 100)
Model.1.vcov <- vcovCR(Model.1, type = "CR2") # Calculate variance-covariance matrix
Model.1.crve <-
  coef_test(Model.1, vcov = Model.1.vcov , test = "Satterthwaite") # Test regression coefficients 
Model.1.crve
```

#### Variance components

```{r}
VarCorr(Model.1, which = "var-cov")
```

#### Standard errors of variances

```{r}
sqrt(diag(varcomp_vcov(Model.1)))
```


## Model 2

#### Prevalence rate change of students identified as having characteristics of dyslexia over seven school years moderated by school district locale (i.e., city, suburban, town, or rural)

```{r, eval = FALSE}
Model.2 <- lme(
    prevalence_rate ~ slope_2017 + slope_change_2018 + slope_change_2019 + 
        slope_change_2020 + slope_change_2021 + slope_change_2022 +
        (slope_2017 + slope_change_2018 + slope_change_2019 + 
             slope_change_2020 + slope_change_2021 + slope_change_2022)*Suburb + 
        (slope_2017 + slope_change_2018 + slope_change_2019 + 
             slope_change_2020 + slope_change_2021 + slope_change_2022)*Town +
        (slope_2017 + slope_change_2018 + slope_change_2019 + 
             slope_change_2020 + slope_change_2021 + slope_change_2022)*Rural, 
    data = data,
    method = "REML",
    random = ~ 1 | district,
    correlation = corAR1(value = 0.2, form = ~ year | district),
    control = list(
        maxIter = 100, 
        msMaxIter = 100, 
        tolerance = 1e-3, 
        opt = "optim",
        optimMethod = "BFGS"
    )
)
```

#### Cluster-robust variance estimation

```{r}
options(width = 100)
Model.2.vcov <- vcovCR(Model.2, type = "CR2") # Calculate variance-covariance matrix
Model.2.crve <-
    coef_test(Model.2, vcov = Model.2.vcov , test = "Satterthwaite") # Test regression coefficients 
Model.2.crve
```

#### Variance components

```{r}
VarCorr(Model.2, which = "var-cov")
```

#### Standard errors of variances

```{r}
sqrt(diag(varcomp_vcov(Model.2)))
```

```{r, echo = FALSE, eval = FALSE}
# save(geom_line_plot, geom_col_plot, Model.0, Model.1, Model.2, file = "output.RData")
```

